New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices

### 2024-10-14 jso.qmd
update smoothness and squintaiblity diagram

Smoothness and squintability are computed following the procedures
outlined in @sec-smoothness and @sec-squintability and as illustrated in
@fig-smoothness and @fig-squintability. To compute smoothness, 500
random bases are simulated. Index values are calculated for each random
basis before fitting the Gaussian process model to obtain the smoothness
measure for the index.

To compute squintability, 50 random bases are sampled and interpolated
to the optimal basis with a step size of 0.005. Index values and
projection distances are calculated for these interpolated bases and the
index values are averaged with a bin width of 0.005. The four-parameter
scaled logistic function (@eq-parametric) is fitted to the index values
against projection distances, estimated by non-linear least squares. The
squintability measure is then calculated as
(@eq-squintability-parametric).

@@

```{r}
#| label: fig-smoothness
#| out.width: 100%
#| fig.cap: "Steps for calculating smoothness in a projection pursuit problem. Given the target shape, data dimension and the index function: (1) sample random bases given the orthonormality contraint and calculate their corresponding index values, and (2) fit a Gaussian process model of index values against the sampled bases to obtain the smoothness measure."
# basis_smoothness <- sample_bases(idx = "splines2d", n_basis = 1000)
# basis_smoothness |>
#    ggplot(aes(x = dist, y = index)) +
#    geom_point() + 
#    labs(x = "Projection distance", y = "Index value") + 
#    theme_bw() +   
#   theme(axis.text = element_text(size = 20), axis.title = element_text(size = 30))
knitr::include_graphics(here::here("figures", "smoothness.png"))
```

```{r}
#| label: fig-squintability
#| out.width: 100%
#| fig.cap: |
#|   Steps for calculating squintability in a projection pursuit problem. Given the target shape, data dimension and the index function: (1) sample random bases given the orthonormality and projection distance contraint, (2) interpolate the sampled bases to the optimal basis to calculate the projection distance and the index value. (3) bin the index values by projection distances to obtain the average index value for each bin, (4) fit the scaled sigmoid function in equation \eqref{eq-squintability} to the binned index values against projection distances using non-linear least square, 5) calculate the squintability measure using equation (@eq-squintability-parametric) with parameters estimated from the model.
# p1 <- sim_sine_6d_spline_projdist |> 
#   ggplot(aes(x = proj_dist, y = index_val)) +
#   geom_point(size = 0.1, alpha = 0.5) + 
#   labs(x = "Projection distance", y = "Index value") + 
#   theme_bw() + 
#   theme(axis.text = element_text(size = 20), axis.title = element_text(size = 30))
# 
# 
# dist_bin <- ceiling(sim_sine_6d_spline_projdist[["proj_dist"]] / 0.1) * 0.1
# dt <- sim_sine_6d_spline_projdist |>
#   dplyr::bind_cols(dist_bin = dist_bin) |>
#   dplyr::group_by(dist_bin) |>
#   dplyr::summarise(index_val = mean(index_val, na.rm = TRUE))
# 
# p2 <-  dt |> 
#   ggplot(aes(x = dist_bin, y = index_val)) + 
#   geom_line() +
#   geom_point() + 
#   labs(x = "Projection distance", y = "Index value") + 
#   theme_bw() + 
#   theme(axis.text = element_text(size = 20), axis.title = element_text(size = 30))
knitr::include_graphics(here::here("figures", "squintability.png"))
```

>>>>

Smoothness and squintability are computed following the procedures
outlined in @sec-smoothness and @sec-squintability and as illustrated in
@fig-smoothness and @fig-squintability. To compute smoothness, 500
random bases are simulated. Index values are calculated for each random
basis, followed by fitting a Gaussian process model to obtain the
smoothness measure for the index.

To compute squintability, 50 random bases are sampled and interpolated
to the optimal basis with a step size of 0.005. Index values and
projection distances are calculated for these interpolated bases. A
binning procedure is applied to average the index values within each
0.005 projection distance bin. The four-parameter scaled logistic
function (@eq-parametric) is fitted to the index values against
projection distances, estimated by non-linear least squares to obtain
the squintability measure as Equation @eq-squintability-parametric.

@@

```{r}
#| label: fig-smoothness
#| out.width: 100%
#| fig.cap: "Steps for calculating smoothness in a projection pursuit problem. Given the target shape, data dimension and the index function: 1) sample random bases given the orthonormality contraint and calculate their corresponding index values, and 2) fit a Gaussian process model of index values against the sampled bases to obtain the smoothness measure."
# basis_smoothness <- sample_bases(idx = "splines2d", n_basis = 1000)
# basis_smoothness |>
#    ggplot(aes(x = dist, y = index)) +
#    geom_point() + 
#    labs(x = "Projection distance", y = "Index value") + 
#    theme_bw() +   
#   theme(axis.text = element_text(size = 20), axis.title = element_text(size = 30))
knitr::include_graphics(here::here("figures", "smoothness.png"))
```

```{r}
#| label: fig-squintability
#| out.width: 100%
#| fig.cap: |
#|   Steps for calculating squintability in a projection pursuit problem. Given the target shape, data dimension and the index function: 1) sample random bases given the orthonormality and projection distance contraint and calculates their corresponding index values, 2) interpolate the sampled bases to the optimal basis and calculate the projection distance and the index value. 3) bin the index values by projection distances through averaging index value, 4) fit the scaled sigmoid function in equation \eqref{eq-squintability} to the binned index values against projection distances using non-linear least square to obtian the squintability measure using equation (@eq-squintability-parametric).
# p1 <- sim_sine_6d_spline_projdist |> 
#   ggplot(aes(x = proj_dist, y = index_val)) +
#   geom_point(size = 0.1, alpha = 0.5) + 
#   labs(x = "Projection distance", y = "Index value") + 
#   theme_bw() + 
#   theme(axis.text = element_text(size = 20), axis.title = element_text(size = 30))
# 
# 
# dist_bin <- ceiling(sim_sine_6d_spline_projdist[["proj_dist"]] / 0.1) * 0.1
# dt <- sim_sine_6d_spline_projdist |>
#   dplyr::bind_cols(dist_bin = dist_bin) |>
#   dplyr::group_by(dist_bin) |>
#   dplyr::summarise(index_val = mean(index_val, na.rm = TRUE))
# 
# p2 <-  dt |> 
#   ggplot(aes(x = dist_bin, y = index_val)) + 
#   geom_line() +
#   geom_point() + 
#   labs(x = "Projection distance", y = "Index value") + 
#   theme_bw() + 
#   theme(axis.text = element_text(size = 20), axis.title = element_text(size = 30))
knitr::include_graphics(here::here("figures", "squintability.png"))
```

### 2024-10-14 jso.tex
update smoothness and squintaiblity diagram

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, title={Algorithm: Jellyfish Optimizer Pseudo Code}, breakable, bottomrule=.15mm, toptitle=1mm, titlerule=0mm, opacityback=0, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, coltitle=black, toprule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6, rightrule=.15mm]

@@

This section describes the details of two simulation studies: one
compares the JSO performance with an existing optimiser, Creeping Random
Search (CRS) (Zhang et al. 2021; Laa and Cook 2020) to explore JSO's
behaviour under different data dimensions. The second simulation first
examines the performance JSO under different JSO hyper-parameters
(number of jellyfish and the maximum number of tries), and then studies
the effect of index properties (smoothness and squintability), along
with JSO hyper-parameters, and data dimension, on the performance of JSO
through a logistic regression.

@@

Smoothness and squintability are computed following the procedures
outlined in Section~\ref{sec-smoothness} and
Section~\ref{sec-squintability} and as illustrated in
Figure~\ref{fig-smoothness} and Figure~\ref{fig-squintability}. To
compute smoothness, 500 random bases are simulated. Index values are
calculated for each random basis before fitting the Gaussian process
model to obtain the smoothness measure for the index.

To compute squintability, 50 random bases are sampled and interpolated
to the optimal basis with a step size of 0.005. Index values and
projection distances are calculated for these interpolated bases and the
index values are averaged with a bin width of 0.005. The four-parameter
scaled logistic function (\ref{eq-parametric}) is fitted to the index
values against projection distances, estimated by non-linear least
squares. The squintability measure is then calculated as
(\ref{eq-squintability-parametric}).

@@

\caption{\label{fig-smoothness}Steps for calculating smoothness in a
projection pursuit problem. Given the target shape, data dimension and
the index function: (1) sample random bases given the orthonormality
contraint and calculate their corresponding index values, and (2) fit a
Gaussian process model of index values against the sampled bases to
obtain the smoothness measure.}

@@

\caption{\label{fig-squintability}Steps for calculating squintability in
a projection pursuit problem. Given the target shape, data dimension and
the index function: (1) sample random bases given the orthonormality and
projection distance contraint, (2) interpolate the sampled bases to the
optimal basis to calculate the projection distance and the index value.
(3) bin the index values by projection distances to obtain the average
index value for each bin, (4) fit the scaled sigmoid function in
equation \eqref{eq-squintability} to the binned index values against
projection distances using non-linear least square, 5) calculate the
squintability measure using equation (\ref{eq-squintability-parametric})
with parameters estimated from the model.}

>>>>

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, toprule=.15mm, colframe=quarto-callout-note-color-frame, breakable, toptitle=1mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, coltitle=black, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, opacitybacktitle=0.6, bottomrule=.15mm, titlerule=0mm, rightrule=.15mm, arc=.35mm, bottomtitle=1mm, left=2mm]

@@

This section describes the details of two simulation studies: one
compares the JSO performance with an existing optimiser, Creeping Random
Search (CRS) (Zhang et al. 2021; Laa and Cook 2020) to explore JSO's
behaviour under different data dimensions. The second simulation first
examines the JSO performance under different hyper-parameters
combinations (number of jellyfish and the maximum number of tries), and
then studies the effect of index properties (smoothness and
squintability), along with JSO hyper-parameters, and data dimension, on
the JSO performance through a logistic regression.

@@

Smoothness and squintability are computed following the procedures
outlined in Section~\ref{sec-smoothness} and
Section~\ref{sec-squintability} and as illustrated in
Figure~\ref{fig-smoothness} and Figure~\ref{fig-squintability}. To
compute smoothness, 500 random bases are simulated. Index values are
calculated for each random basis, followed by fitting a Gaussian process
model to obtain the smoothness measure for the index.

To compute squintability, 50 random bases are sampled and interpolated
to the optimal basis with a step size of 0.005. Index values and
projection distances are calculated for these interpolated bases. A
binning procedure is applied to average the index values within each
0.005 projection distance bin. The four-parameter scaled logistic
function (\ref{eq-parametric}) is fitted to the index values against
projection distances, estimated by non-linear least squares to obtain
the squintability measure as Equation \ref{eq-squintability-parametric}.

@@

\caption{\label{fig-smoothness}Steps for calculating smoothness in a
projection pursuit problem. Given the target shape, data dimension and
the index function: 1) sample random bases given the orthonormality
contraint and calculate their corresponding index values, and 2) fit a
Gaussian process model of index values against the sampled bases to
obtain the smoothness measure.}

@@

\caption{\label{fig-squintability}Steps for calculating squintability in
a projection pursuit problem. Given the target shape, data dimension and
the index function: 1) sample random bases given the orthonormality and
projection distance contraint and calculates their corresponding index
values, 2) interpolate the sampled bases to the optimal basis and
calculate the projection distance and the index value. 3) bin the index
values by projection distances through averaging index value, 4) fit the
scaled sigmoid function in equation \eqref{eq-squintability} to the
binned index values against projection distances using non-linear least
square to obtian the squintability measure using equation
(\ref{eq-squintability-parametric}).}
